\documentclass[10pt,landscape,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{tikz}
\usetikzlibrary{shapes,positioning,arrows,fit,calc,graphs,graphs.standard}
%\usepackage[nosf]{kpfonts}
%\usepackage[t1]{sourcesanspro}
\usepackage{multicol}
\usepackage{wrapfig}
\usepackage[top=10mm,bottom=10mm,left=10mm,right=10mm]{geometry}
\usepackage[framemethod=tikz]{mdframed}
%\usepackage{microtype}
\usepackage{pdfpages}

\usepackage[pdftex,
            pdfauthor={bandicode},
            pdftitle={C++ Cheatsheet - Memory Management},
            pdfsubject={Memory Management},
            pdfkeywords={memory,smart pointers,allocators},
            pdfproducer={Latex},
            pdfcreator={pdflatex}]{hyperref}

\include{common/def}

\begin{document}

\CheatsheetTitle{Memory Management}

\small
\begin{multicols*}{3}

\section*{Manual memory management}

The keywords \tcode{\href{https://en.cppreference.com/w/cpp/language/new}{new}} and 
\tcode{\href{https://en.cppreference.com/w/cpp/language/delete}{delete}} can be used 
to create and destroy object on the heap.

\begin{codeblock}
Foo *foo = new Foo;
delete foo;
\end{codeblock}

Unlike stack-allocated arrays, arrays allocated on the heap can have sizes not known 
at compile time. Memory allocated with \tfcode{new[]} should be freed with \tfcode{delete[]}.

\begin{codeblock}
int *vec = new int[size];
delete[] vec;
\end{codeblock}

\headline{Initialization of dynamically allocated objects}

Objects allocated with \tfcode{new} can be initialized.

\begin{codeblock}
int *value = new int(42);
double *vec = new double[]{1.0, 2.0, 3.0};
\end{codeblock}

\subsection*{Placement new}

An object can be constructed in an already allocated storage using placement new.
This is done by providing the memory buffer to \tfcode{new}. 
Note that in such situation, the object should be destroyed by calling its destructor 
explicitly. 

\begin{codeblock}
char* ptr = new char[sizeof(T)];
T* tptr = new (ptr) T;
tptr->~T();
delete[] ptr;
\end{codeblock}

\section*{Automatic memory management}

Today's preferred way to deal with dynamically-allocated memory is 
through smart-pointers defined in header \tcode{<memory>}.

\subsection*{Smart pointers}

\bgroup
\def\arraystretch{1.5}
\def\cellwidth{6cm}
\begin{tabular}{ l L{\cellwidth} } 
 \arrayrulecolor{gray}\hline
   \hlinkb{unique-ptr-anchor}{\tcode{unique\_ptr}} & 
   smart pointer with unique object ownership semantics \tmark{class template} \\ 
 \hline
   \hlinkb{shared-ptr-anchor}{\tcode{shared\_ptr}} & 
   smart pointer with shared object ownership semantics \tmark{class template} \\ 
 \hline
   \hlinkb{weak-ptr-anchor}{\tcode{weak\_ptr}} & 
   weak reference to an object managed by \hlinkb{shared-ptr-anchor}{\tcode{shared\_ptr}} \tmark{class template} \\ 
\end{tabular}
\egroup

\subsubsection*{\hypertarget{unique-ptr-anchor}{\tcode{unique\_ptr}}}

The class template \hrefcode{https://en.cppreference.com/w/cpp/memory/unique_ptr}{std::unique\_ptr} 
owns an object that it destroys when the unique-pointer goes out of scope. 
This class satisfies the requirements \hrefrequirements{https://en.cppreference.com/w/cpp/named_req/MoveConstructible}{MoveConstructible} 
and \hrefrequirements{https://en.cppreference.com/w/cpp/named_req/MoveAssignable}{MoveAssignable} but it is neither 
copy constructible nor copyable. 

\bgroup
\def\arraystretch{0}
\def\cellwidth{7cm}
\begin{tabular}{ L{\cellwidth} l } 
 \arrayrulecolor{gray}\hline
   \begin{hsynopsis}
template<
    class T,
    class Deleter = std::default_delete<T>
> class unique_ptr;
   \end{hsynopsis}   & 
   (1) \\ 
\end{tabular}
\egroup

\vspace{-3ex}

The class overloads \tcode{operator*} and \tcode{operator->} so that it can 
be used as if it was a pointer to the underlying type.

\begin{codeblock}
{
  int* raw_ptr = new int(0);
  std::unique_ptr<int> ptr{raw_ptr};
  *ptr = 42;
  // the integer is destroyed by unique\_ptr at
  // the end of this block
}
\end{codeblock}

The class also provides a partial specialization for array types.

\begin{codeblock}
std::unique_ptr<int[]> ptr{new int[7]};
ptr[0] = 0;
\end{codeblock}

\headline{Essentials}

\bgroup
\def\arraystretch{1.5}
\def\cellwidth{6cm}
\begin{tabular}{ l L{\cellwidth} } 
 \arrayrulecolor{gray}\hline
    \hrefbf{https://en.cppreference.com/w/cpp/memory/unique_ptr/get}{\tcode{get}} & 
   returns a pointer to the managed object \tmark{public member function} \\ 
 \hline
   \hrefbf{https://en.cppreference.com/w/cpp/memory/unique_ptr/release}{\tcode{release}} & 
   returns a pointer to the managed object and releases the ownership  \tmark{public member function} \\ 
 \hline
   \hrefbf{https://en.cppreference.com/w/cpp/memory/unique_ptr/reset}{\tcode{reset}} & 
   replaces the managed object  \tmark{public member function} \\ 
\end{tabular}
\egroup

\subsubsection*{\hypertarget{shared-ptr-anchor}{\tcode{shared\_ptr}}}

The class template \hrefcode{https://en.cppreference.com/w/cpp/memory/shared_ptr}{std::shared\_ptr} 
provides a reference-counted pointer. 
When a shared pointer is copied, the reference counter is incremented. 
The managed object is destroyed when the last shared pointer referencing it is destroyed.

\bgroup
\def\arraystretch{0}
\def\cellwidth{8cm}
\begin{tabular}{ L{\cellwidth} l } 
 \arrayrulecolor{gray}\hline
   \begin{hsynopsis}
template<class T> class shared_ptr;
   \end{hsynopsis}   & \\ 
\end{tabular}
\egroup

\headline{Essentials}

\bgroup
\def\arraystretch{1.5}
\def\cellwidth{6cm}
\begin{tabular}{ l L{\cellwidth} } 
 \arrayrulecolor{gray}\hline
    \hrefbf{https://en.cppreference.com/w/cpp/memory/shared_ptr/get}{\tcode{get}} & 
   returns a pointer to the managed object \tmark{public member function} \\ 
 \hline
   \hrefbf{https://en.cppreference.com/w/cpp/memory/shared_ptr/reset}{\tcode{reset}} & 
   replaces the managed object \tmark{public member function} \\ 
 \hline
   \hrefbf{https://en.cppreference.com/w/cpp/memory/shared_ptr/use_count}{\tcode{use\_count}} & 
   returns the number of shared\_ptr objects referring to the same managed object  \tmark{public member function} \\ 
\end{tabular}
\egroup

Typical implementations contain a pointer to the managed object 
and a pointer to a \emph{control-block} that stores the number of 
shared pointers that own the object and the number of weak pointers 
that refer to the object.

\subsubsection*{\hypertarget{weak-ptr-anchor}{\tcode{weak\_ptr}}}

Weak pointers can be constructed from shared pointers to create a non-owning reference 
to a managed object.
Weak pointers can be used when temporary ownership is sufficient, or to 
break reference cycles.

\headline{Essentials}

\bgroup
\def\arraystretch{1.5}
\def\cellwidth{6cm}
\begin{tabular}{ l L{\cellwidth} } 
 \arrayrulecolor{gray}\hline
    \hrefbf{https://en.cppreference.com/w/cpp/memory/weak_ptr/expired}{\tcode{expired}} & 
   checks whether the referenced object was destroyed \tmark{public member function} \\ 
 \hline
   \hrefbf{https://en.cppreference.com/w/cpp/memory/weak_ptr/lock}{\tcode{lock}} & 
   creates a shared\_ptr that manages the referenced object \tmark{public member function} \\ 
\end{tabular}
\egroup

\headline{Use case: breaking reference cycles}

\begin{codeblock}
struct Node
{
  std::vector<std::shared_ptr<Node>> children;
  std::weak_ptr<Node> parent;

  Node(const std::shared_ptr<Node>& p = nullptr) 
    : parent(p) { }
  virtual ~Node() = default;
};
\end{codeblock}

\subsection*{Helper functions}

\bgroup
\def\arraystretch{1.5}
\def\cellwidth{6cm}
\begin{tabular}{ l L{\cellwidth} } 
 \arrayrulecolor{gray}\hline
   \hrefbf{https://en.cppreference.com/w/cpp/memory/shared_ptr/make_shared}{\tcode{make\_shared}} & 
   constructs an wrap an object in a shared pointer \tmark{function template} \\ 
 \hline
   \hrefbf{https://en.cppreference.com/w/cpp/memory/unique_ptr/make_unique}{\tcode{make\_unique}} & 
   constructs an wrap an object in a unique pointer \tmark{function template} \tsincecxx{14} \\ 
\end{tabular}
\egroup

When using \tcode{make\_shared}, only a single memory block large enough to store 
the managed object and the control block is allocated (the managed object is constructed in-place). 
Pros: you save a dynamic memory allocation compared to using \tcode{shared\_ptr} constructor; 
Cons: the allocated memory is released when the last weak pointer referencing the object 
is destroyed.

\subsection*{Helper class \tcode{enable\_shared\_from\_this}}

The \hrefcode{https://en.cppreference.com/w/cpp/memory/enable_shared_from_this/shared_from_this}{shared\_from\_this} 
function can be used to obtain a shared pointer from a raw pointer.
To enable this feature, your class must derive from
\hrefcode{https://en.cppreference.com/w/cpp/memory/enable_shared_from_this}{std::enable\_shared\_from\_this}.

\bgroup
\def\arraystretch{0}
\def\cellwidth{8cm}
\begin{tabular}{ L{\cellwidth} l } 
 \arrayrulecolor{gray}\hline
   \begin{hsynopsis}
template<class T> class enable_shared_from_this;
   \end{hsynopsis}   & \\ 
\end{tabular}
\egroup

Note that classes deriving from \tcode{enable\_shared\_from\_this} shouldn't be constructed on the stack.

\headline{Example:}

\begin{codeblock}
class Foo : 
  public std::enable_shared_from_this<Foo> 
{ 

};
/* ... */
Foo* foo = ...;
std::shared_ptr<Foo> ptr = foo->shared_from_this();
\end{codeblock}

\section*{Allocators}

Standard library containers use an \hrefrequirements{https://en.cppreference.com/w/cpp/named_req/Allocator}{Allocator} 
to allocate and release memory. 
The default allocator,
\hrefcode{https://en.cppreference.com/w/cpp/memory/allocator}{std::allocator}, is used 
if you don't specify any (and is sufficient most of the time).

\bgroup
\def\arraystretch{0}
\def\cellwidth{8cm}
\begin{tabular}{ L{\cellwidth} l } 
 \arrayrulecolor{gray}\hline
   \begin{hsynopsis}
template<class T>  struct allocator;
   \end{hsynopsis}   & \\ 
\end{tabular}
\egroup

\vspace{-3ex}

Custom allocators should at least provide the following:

\headline{Types}

\bgroup
\def\arraystretch{1.5}
\def\cellwidth{3cm}
\begin{tabular}{ L{\cellwidth} L{\cellwidth} } 
  \arrayrulecolor{gray}
  \hline
    \textbf{Type} & 
    \textbf{Definition} \\
  \hline
    \tcode{value\_type} & 
    \tfcode{T} \\
\end{tabular}
\egroup

\headline{Member functions}

\bgroup
\def\arraystretch{1.5}
\def\cellwidth{6cm}
\begin{tabular}{ l L{\cellwidth} } 
  \arrayrulecolor{gray}
  \hline    
    \hrefbf{https://en.cppreference.com/w/cpp/memory/allocator/allocate}{\tcode{allocate}} & 
    allocates uninitialized storage \tmark{public member function} \\ 
  \hline
    \hrefbf{https://en.cppreference.com/w/cpp/memory/allocator/deallocate}{\tcode{deallocate}} & 
    deallocates storage \tmark{public member function} \\ 
\end{tabular}
\egroup

\headline{Minimal allocator with debug output:}

\begin{codeblock}
template<class Tp>
struct Alloc {
  typedef Tp value_type;
  Alloc() = default;
  
  template<class T> Alloc(const Alloc<T>&) {}
  Tp* allocate(std::size_t n) 
  {
    n *= sizeof(Tp);
    std::cout << "allocating " << n << " bytes\n";
    return static_cast<Tp*>(::operator new(n));
  }
  
  void deallocate(Tp* p, std::size_t n) 
  {
    n *= sizeof(Tp);
    std::cout << "deallocating " << n << " bytes\n";
    ::operator delete(p);
  }
};
\end{codeblock}

\begin{codeblock}
template <class T, class U>
bool operator==(const Alloc<T>&, const Alloc<U>&) 
{ return true; }

template <class T, class U>
bool operator!=(const Alloc<T>&, const Alloc<U>&) 
{ return false; }
\end{codeblock}

You can specify various properties of your allocator by specializing the class template 
\hrefbf{https://en.cppreference.com/w/cpp/memory/allocator_traits}{\tcode{std::allocator\_traits}} 
for your allocator.

\section*{Overloading operator new \& delete}

You can track allocation and deallocation of user-defined types, or 
more generally customize the allocation process, by overloading
\hrefbf{https://en.cppreference.com/w/cpp/memory/new/operator_new}{\tcode{operator new}}.
This operator supports both global replacements and class-specific overloads.

These overloads are responsible for allocating and deallocating bytes of memory 
on which actual object are constructed.

\begin{codeblock}
struct X 
{
  static void* operator new(std::size_t sz) 
  {
    std::cout << "Allocating X (" << sz << " bytes)\n";
    return ::operator new(sz);
  }

  static void operator delete(void* ptr)
  {
    std::cout << "Deallocating X\n";
    ::operator delete(ptr);
  }
};
\end{codeblock}

\section*{Additional Resources}

\href{https://isocpp.org/wiki/faq/freestore-mgmt}{Memory Management}, \href{https://isocpp.org/faq}{C++ Super-FAQ}, \href{https://isocpp.org}{isocpp.org}.

\href{https://en.cppreference.com/w/cpp/memory/new}{Low level memory management}, \href{https://en.cppreference.com/}{en.cppreference.com}.

\vspace{40ex}

\section*{Notes}

\begin{userendnotes}
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
  \hgrayrule
\end{userendnotes}

\end{multicols*}

\end{document}